Correcting for bias in high throughput sequencing data

### 2011-06-15 seqbias-paper.tex
more revisions

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
\caption{Blah blah blah.}
\label{fig:models}
\end{figure*}

@@

%% leaving off here for now
%% --------------------------------------------------------

@@

Plotting the nucleotide frequencies (Figure  \ref{fig:freqs}) we observe an
obvious bias. To quantify the non-uniformity observed in these plots, we use the
symmetrized Kullback-Leibler (KL) divergence \cite{Kullback1951}. Comparing only
nucleotide frequencies may miss dinucleotide or higher order bias. Thus, for a fair
comparison, we compute the KL divergence for $k$-mers, with $k \in \{ 1 \dots 6
\}$.

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as,
$$D( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
Where the sum is over all $k$-mers. We compute the divergence at each
position relative to the read start, by tallying (overlapping) $k$-mer
frequencies at each position, weighted by the bias coefficient at that position,
and comparing to a background distribution. The background distribution is
estimated by averaging the $k$-mer frequencies at positions $-60 \dots -20$ and
$20 \dots 60$, where the observed bias was very low in each data set.  We further
take the precaution of ignoring any positions with more than 10,000 reads, to
avoid the measure being dominated by any small group of outliers. The only
qualitative effect this was seen to have was in the Katze data set, in which the
frequency plots are otherwise quite skewed.

The result of this this analysis is plotted in Figure
\ref{fig:kl} for the Trapnell data set. Similar plots for the other data sets are
provided in the supplement.

\begin{figure*}
\centerline{\includegraphics[width=0.9\textwidth]{fig/kl.eps}}
\caption{Plots of the KL divergence over overlapping $k$-mers in the Trapnell
experiment.}
\label{fig:kl}
\end{figure*}

To summarize these plots, we summed the divergence from position -20
through 20, for each of the methods, then normalized this number by dividing by
the summed divergence before adjustment. The results of this analysis are plotted in
Figure \ref{fig:kl-delta}.

%% TODO
\begin{comment}
\begin{figurehere}
\begin{center}
\includegraphics[width=0.35\textwidth]{fig/kl-delta.pdf}
\end{center}
\caption{Summed KL divergence, normalized by unadjusted divergence, so that a
number less than 1 indicates a decrease in divergence, or an increase in
uniformity.}
\label{fig:kl-delta}
\end{figurehere}
\end{comment}

@@

In this comparison, we measure how well the counts conform to a Poisson process.
The assumption of positional read counts following a Poisson distribution has
been shown to be a poor fit \cite{Srivastava2010}, yet measuring the improvement in
fit derived from correcting for technical bias remains a principled and easily
interpreted criteria.

We perform maximum-likelihood fitting of two models. In the null model, the
Poisson rate is fixed for each exon. That is, for position $j$ within
gene $i$, the rate is,
$$ \lambda_{ij} = a_i $$
Where $a_i$ is the parameter being fit. For comparison, we then fit a model in
which the rate is proportional to the predicted bias coefficients.
$$ \lambda'_{ij} = a_i b_{ij} $$

If the null model has log-likelihood $L$ and the bias corrected model
$L'$, we measure the change in fit as,
$$ \log \left( \frac{ L_{i} } { L'_{i} } \right) $$
When the fit improves, this number is positive, and negative when it is
worsened. The values computed from our test are listed in Table \ref{tab:pois}.

\begin{comment}
\begin{tablehere}
\begin{center}
\tiny{
\begin{tabular}{lrrrrrr}
         &\textbf{Avg-7mer}&\textbf{7mer}&\textbf{4mer}&\textbf{GLM}&\textbf{MART}&\textbf{BN} \\
\textbf{Katze}            &  0.004  & 0.017  & 0.055 &0.166  &0.248&0.305 \\
\textbf{Wetterbom}        & -0.076  &-0.058  &-0.002 &0.240  &0.070&0.557 \\
\textbf{Bullard}          &  0.124  & 0.192  & 0.139 &0.285  &0.198&0.380 \\
\textbf{Mortazavi}        &  0.059  & 0.123  & 0.099 &0.284  &0.310&0.401 \\
\textbf{Trapnell}         &  0.094  & 0.163  & 0.104 &0.265  &0.332&0.349 \\
\end{tabular}
}
\end{center}
\caption{Change in fit, compared to the null model of uniform Poisson rate across exons.}
\label{tab:pois}
\end{tablehere}
\end{comment}

Computing one number to summarize the fit of the model ignores variance in the
likelihood from exon to exon. This number could be dominated by a few test
regions. Figure \ref{fig:pois} plots the distribution of the change in fit
across the 1000 test exons, revealing a similar, but more nuanced result.

%% TODO
\begin{figure}
\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
\caption{Plotted is the change in fit of each of the test exons. A
positive value indicates a better fit in the bias corrected model.}
\label{fig:pois}
\end{figure}

@@

In our results, estimating initial 4-mer or 7-mer frequencies was not seen to be
as effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate 7-mer
frequencies, a likely explanation is that the model is overfit to the training
set. Though the training set consisted of at least 600,000 reads in
each experiment, it seems more reads, or a greater diversity of reads are needed
to produce an accurate estimate.

Our method generalizes this approach, attempting to overcome this problem by
using an estimation of sequence probability that is more robust to overfitting
and can account for bias beyond the initial heptamer. This approach is
competitive with that of Li, et. al., despite not requiring gene annotations.

Because we do not require annotations, CHiP-Seq, and other high-throughput
sequencing experiments, may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one CHiP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, supplementary). Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in CHiP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into CHiP-Seq peak calling algorithms, and potentially improve
accuracy.

Our method can also very naturally capture dependency between nucleotides in the
sequence. This of particular importance when working with reads generated from
the AB SOLiD platform, in which the di-base, ``colorspace'' coding scheme is
used, encoding two adjacent nucleotides as a single color. A color bias at one
position will manifest itself as a nucleotide bias at two positions. Capturing
this effect requires a model that considers the dependency between nearby
positions. In our results, the MART model, as well as our own, perform
significantly better than the GLM model, illustrating this proposition.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see,
supplementary). Similarly, in Figure \ref{fig:freqs}, the three data sets
sequenced on the Illumina platform display similar patterns of nonuniformity,
yet differ in magnitude. Batch effects in RNA-Seq remain a legitimate concern,
that should be evaluated and corrected for.

>>>>

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
%% TODO: caption
\caption{Blah blah blah.}
\label{fig:models}
\end{figure*}

@@

@@

Plotting the nucleotide frequencies (Figure  \ref{fig:freqs}) we observe an
obvious bias. To quantify the non-uniformity observed in these plots, we use the

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as, symmetrized Kullback-Leibler (KL)
divergence \citep{Kullback1951}.
%Comparing only nucleotide frequencies may miss
%dinucleotide or higher order bias. Thus, for a fair comparison, we compute the
%KL divergence for $k$-mers, with $k \in \{ 1 \dots 6 \}$.
$$D_k( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
where the sum is over all $k$-mers.

%% --------------------

To avoid the risk of a small number of reads with many duplicates dominating the
measure, we do not count duplicate reads when computing the KL divergence. Since we
explicitly chose exons with high coverage, not counting duplicates may not
fully capturing the effects of the bias correction methods. Thus, for each exon,
we compute the number of duplicates for each read, and count only reads in the
upper half. Under the assumption of uniform sampling, the set of reads falling
in the upper half should not depend on sequence, and we should expect the KL
divergence to be low.

We compute the divergence by reweighting the read counts using the predicted
bias coefficient, choosing the upper half in each exon, discarding duplicate
reads, and then tallying frequencies of overlapping $k$-mers. The $k$-mer
distribution obtained are then compared to a background distribution obtained by
redistributing reads uniformly at random within their exons.

We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
remaining cases are plotted in Supplementary Section 3.
%% TODO: what section, for real?

\begin{figure*}
\centerline{\includegraphics[width=0.9\textwidth]{fig/kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frquencies expected under the assumption of unifom sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted reads counts as well as after adjusting reads counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

In this comparison, we measure how well the counts conform to a Poisson process.
The assumption of positional read counts following a Poisson distribution is
known to be a poor fit \citep{Srivastava2010}, but measuring the improvement in
the fit derived from correcting for bias remains a principled and easily
interpreted criterion.

We perform maximum-likelihood fitting of two models. In the null model, the
Poisson rate is fixed for each exon. That is, for position $j$ within
exon $i$, the rate is
$$ \lambda_{ij} = a_i $$
where $a_i$ is the parameter being fit. For comparison, we then fit a model in
which the rate is also proportional to the predicted bias coefficients:
$$ \lambda'_{ij} = a_i b_{ij} $$

If the null model has log-likelihood $L$, and the bias-corrected model $L'$, a
simple goodness of fit measure is the improvement in log-likelihood (a statistic
commonly known as McFadden's pesudo-coefficient of determination
\citep{McFadden1974}), defined as,
$$R^2 = 1 - \frac{L'}{L}$$

This measure can be interpreted as the improvement in fit over the null model,
with $R^2 = 1$ indicating a prefect fit, and a negitve numbers, an increasingly
worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
proportion of variability eplained by the model.

We compue $R^2$ for each of the test exons, giving us a sense of the variability
of the effectiveness of each model. The results of this analysis are plotted in
Figure \ref{fig:pois}.  To summarize each model with a single number, we can
examine the median $R^2$ value, as listed in Table \ref{tab:pois}.

\begin{figure}
\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we measure a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive when uniformity is increased,
and negative when decreased. Boxes are plotted to mark the 25\%, 50\%, and 75\%
quantiles, with whiskers extending to 1.5 times the inter-quartile range (i.e.,
the span between tho 25\% and 75\% quantiles).
}

    \label{fig:pois}
\end{figure}

\begin{table}
\processtable{
    The $R^2$ goodness of fit statistic measures increased uniformity in read
    coverage, after correcting for bias. Here the median $R^2$ across the test
    exons is listed for each method and simple. A positive $R^2$ indicates a
    better fit.\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & 0.168 & 0.016 & 0.066 & -0.079 \\
Katze     & 0.283 & 0.243 & 0.158 & 0.033 \\
Bullard   & 0.271 & 0.163 & 0.224 & 0.157 \\
Mortazavi & 0.225 & 0.210 & 0.197 & 0.091 \\
Trapnell  & 0.298 & 0.289 & 0.248 & 0.138
    \end{tabular}
}{}
\end{table}

@@

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate
heptamer frequencies, a likely explanation is that the model is overfit to the
training set. Though the training set consisted of at least 600,000 reads in
each experiment, it seems more reads, or a greater diversity of reads are needed
to produce an accurate or useful estimate.

Our method generalizes this approach, attempting to overcome this problem by
using an estimation of sequence probability that is more robust to overfitting
and can account for bias beyond the initial heptamer. In all our tests, this
approach was at least as effective as those of \citet{Li2010}, despite not
requiring gene annotations, or manual selection of training examples.

We have withheld any direct comparison to the method described by
\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
this method is superficially similar to our own, a proper comparison is
difficult, as the software can not be applied independently of estimating
transcript abundance in FPKM using Cufflinks. Fairness would dictate that
competing methods be substituted in FPKM estimation, or that a seperate
interface be written to the Cufflinks bias correction method---both comparison
requiring significant effort.

Though the Cufflinks method and our own both use graphical models to estimate
sequence probabilities, we make no restriction on the graph other than
acyclicity, and go to considerable effort to efficiently approximate the optimal
structure for each data set, rathor than using a fixed structure, as in
Cufflinks. A ``one size fits all'' approach likely works quite well in many
cases, yet the observed specificity of the bias to protocal and platform argues
against it. For example, the structures learned by our method (Figure
\ref{fig:models}), are considerably different between those sequenced on an
Illumina platform versus an AB platform.

%% TODO: discussion of run-time, etc

Because we do not require annotations, CHiP-Seq, and other high-throughput
sequencing experiments may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one CHiP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, Supplementary Section 4).
%% TODO: which actual section ?
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in CHiP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into CHiP-Seq peak calling algorithms, and potentially improve
accuracy.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see,
Supplementary Section 5).
%% TODO: which actual section ?
Similarly, in Figure \ref{fig:freqs}, the three data sets
sequenced on the Illumina platform display similar patterns of nonuniformity,
yet differ in magnitude. Batch effects in RNA-Seq remain a legitimate concern,
that should be evaluated and corrected for.
